%! Author = ferran
%! Date = 14/10/23


% Preamble
\documentclass[11pt]{article}

\usepackage[margin=3cm]{geometry}

% Packages
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{amsfonts}
% Packages
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{tikz-cd}
\usepackage{wasysym}
\usepackage{nicefrac}

\title{Experiments on the splitting of cyclotomic polynomials}
\author{Ferran Espuña Bertomeu}


\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{claim}[theorem]{Claim}

\theoremstyle{definition}
\newtheorem{defn}[theorem]{Definition}
\newtheorem{rk}[theorem]{Remark}



% Document
\begin{document}

    \maketitle

    \section{$\phi_{11}$ in $\mathbb{F}_p$}

    \noindent In this part I will explain the experiments I have done on the splitting of cyclotomic polynomials in fields of prime order.
    On the first experiment, I have computed how the polynomial $\phi_{11}(X) = X^{10} + X^9 + \cdots + X + 1$ splits over $\mathbb{F}_p$,
    for $p$ a prime number.
    I have done this for all primes $p$ less than $200,000$ (excluding $11$).
    The goal of this experiment is to see if there is a pattern on how many irreducible factors $\phi_{11}$ has
    over $\mathbb{F}_p$.
    Excluding the prime $p = 11$, the polynomial $\phi_{11}$ is separable over $\mathbb{F}_p$ for all primes $p$, so the
    extension $\mathbb{F}_p(\zeta_{11})\mid\mathbb{F}_p$ is Galois, where $\zeta_{11}$ is a primitive $11$-th root of unity.
    Therefore, the polynomial $\phi_{11}$ splits into
    irreducible factors of the same degree.
    The only options are:
    \begin{enumerate}
        \item $\phi_{11}$ is irreducible over $\mathbb{F}_p$.
        \item $\phi_{11}$ splits into two irreducible factors of degree $5$ over $\mathbb{F}_p$.
        \item $\phi_{11}$ splits into five irreducible factors of degree $2$ over $\mathbb{F}_p$.
        \item $\phi_{11}$ splits into ten irreducible factors of degree $1$ over $\mathbb{F}_p$.
    \end{enumerate}

    \noindent Note that the last case corresponds to the case where $\mathbb{F}_p(\zeta_{11}) = \mathbb{F}_p$, that is,
    $\mathbb{F}_p$ contains all the $11$-th roots of unity.
    I distinguish between the cases using the Berlekamp algorithm:
    Every time I find a nontrivial factorization of $\phi_{11}$, I run the algorithm again on the smallest of the two factors,
    until I get a polynomial whose number of irreducible factors I can deduce (an irreducible one, or one of degree not dividing $10$,
    in which case I deduce $\phi_{11}$ splits into degree 1 factors).


    \noindent Initially, I wanted to just know how often each case happens, so I tallied the number of times each case happened in the following table:

    \begin{center}
        \begin{tabular}{|c|c|c|c|c|}
            \hline
            Number of irreducible factors mod p & 1 & 2 & 5 & 10 \\
            \hline
            Number of primes p & 7213 & 7178 & 1798 & 1794 \\
            \hline
        \end{tabular}
    \end{center}

    \noindent There are a total of $17983$ primes less than $200,000$ (excluding $11$). This means that,
    for a random prime $p$ less than a million, the probability of each case happening is approximately:

    \begin{center}
        \begin{tabular}{|c|c|c|c|c|}
            \hline
            Number of irreducible factors mod p & 1 & 2 & 5 & 10 \\
            \hline
            fraction of primes p & 0.401 & 0.399 & 0.100 & 0.100 \\
            \hline
        \end{tabular}
    \end{center}

    \noindent These fractions are very close to fractions with denominator $10$, which is suspicious as $10 = 11 - 1$.
    This prompted me to take a look at the remainders of each prime $p$ modulo $11$.
    Sure enough, the remainders perfectly charaterized the splitting behaviour of $\phi_{11}$ modulo each prime,
    for all primes less than a million.
    In particular: \pagebreak

    \begin{itemize}
        \item The first case happens if and only if $p \equiv 2, 6, 7, 8 \pmod{11}$.
        \item The second case happens if and only if $p \equiv 3, 4, 5, 9 \pmod{11}$.
        \item The third case happens if and only if $p \equiv 10 \pmod{11}$.
        \item The fourth case happens if and only if $p \equiv 1 \pmod{11}$.
    \end{itemize}

    \noindent As it turns out, the \emph{degree} of each irreducible factor is exactly the order of $p$ modulo $11$.
    This is not a coincidence.
    We can prove this the fact that the degree of each irreducible factor is the order of one of its roots (a primitive $11$-th root of unity)
    under the Frobenius automorphism $\varphi: x \mapsto x^p$
    (as we are in a finite extension of $\mathbb{F}_p$). But $1 = \varphi^{k}(\zeta_{11}) = \zeta_{11}^{p^k} \iff p^k \equiv 1 \pmod{11}$,
    so the order of $\zeta_{11}$ under $\varphi$ is exactly the order of $p$ modulo $11$.
    Of course, this generalizes to any $q$-th cyclotomic polynomial for any prime $q$.
    As for the reason why the fractions coincide with the fraction of remainders modulo $11$,
    this is explained by Dirichlet's theorem on arithmetic progressions.

    \noindent Through my experiments, I have also made the following remarkable observation:
    the number of irreducible factors is equal to the
    dimension of the kernel of the relevant matrix,
    at least to the extent of my experiments.
    I have tried to prove this, but I have not been able to do so.
    An open question is whether this is true in general,
    and for arbitrary cyclotomic polynomials (or even Galois extensions?).
    or if it is just a coincidence.

    \section{$[T^2 + T + 1]$ in $\mathbb{F}_2[T] / (\pi)$}

    In this case, we have seen in class that the relevant Galois group is isomorphic to \linebreak
    ${(\mathbb{F}_2[T]/(T^2+T+1))^{\times}}$,
    which is a cyclic group of order $3$.
    The only options are therefore that the polynomial $T^2 + T + 1$ splits completely into linear factors, or it is irreducible.
    $\pi$ of $\mathbb{F}_2[T]$ either
    splits completely into linear factors (whenever $\pi \equiv 1 \pmod{T^2 + T + 1}$), or it is irreducible (whenever $\pi \equiv 1$ or $T \pmod{T^2 + T + 1}$).
    We can generate primes of $\mathbb{F}_2[T]$ by running the Berlekamp algorithm on polynomials of degree up to $16$ and seeing which are irreducible.
    Then we can compile statistics on each case:

    \begin{center}
        \begin{tabular}{|c|c|c|c|}
            \hline
            $\pi \pmod M$  & $1$ & $T$ & $T + 1$ \\
            \hline
             Number of polynomials & 2929 & 2935 & 2935 \\
            \hline
        \end{tabular}
    \end{center}

    \noindent We see that the number of polynomials in each case is very close to $\frac{1}{3}$ of the total number of polynomials.
    I hypotesize that this is actually the case when we make our maximum degree go to infinity.
    In that case, the polynomial splits into linear factors if and only if $\pi \equiv 1 \pmod{T^2 + T + 1}$, that is,
    one third of the time.
    Meanwhile, $\pi$ is irreducible if and only if $\pi \equiv 1$ or $T \pmod{T^2 + T + 1}$, that is,
    two thirds of the time.

\end{document}

%Counter({(1, 6): 7876, (1, 7): 7874, (1, 8): 7873, (10, 1): 7858, (2, 5): 7853, (1, 2): 7843, (2, 4): 7839, (5, 10): 7839, (2, 9): 7828, (2, 3): 7814})
%Counter({1: 31466, 2: 31334, 10: 7858, 5: 7839})

% Up to 16: {'x': 2935, 'x + 1': 2935, '1': 2929, '0': 1}

%Up to 10000
%Counter({(2, 9): 126, (1, 6): 125, (10, 1): 125, (1, 7): 124, (1, 8): 124, (5, 10): 123, (2, 3): 122, (2, 5): 121, (1, 2): 120, (2, 4): 118})
%Counter({1: 493, 2: 487, 10: 125, 5: 123})

%Up to 200000
%Counter({(1, 2): 1816, (1, 8): 1804, (2, 3): 1803, (1, 7): 1801, (2, 5): 1799, (5, 10): 1798, (10, 1): 1794, (1, 6): 1792, (2, 9): 1792, (2, 4): 1784})
%Counter({1: 7213, 2: 7178, 5: 1798, 10: 1794})